%!TEX encoding = UTF-8 Unicode

\addchap{Acknowledgments}

Experiments presented in this paper were carried out using the \mbox{Grid'5000}
testbed, supported by a scientific interest group hosted by \mbox{Inria} and including
\mbox{CNRS}, \mbox{RENATER} and several Universities as well as other organizations.

I would like to thank all the authors who upload their papers on open archives like HAL or Arxiv. Likewise, thanks a lot
to Alexandra \textsc{Elbakyan} for the creation of Sci-Hub, which helped me a lot for retrieving articles behind
paywalls.

Thank you to the members of the jury for kindly accepting to be part of this committee. In particular, thanks a lot to
the two reviewers, Patrick \textsc{Carribault} and Henri \textsc{Casanova}, who made excellent comments on the
manuscript which helped to improve the final version.

Thank you to all the Simgrid contributors for creating and maintaining the software on which is based a great part of
this thesis. They have been very helpful and friendly in numerous occasions.

Thank you to the Grid'5000 staff, I cannot emphasize enough how much their work is important. This platform is a
wonderful tool for computer science experiments.

Thanks a lot to the members of Polamove and Dataris teams who welcomed me for these last years. The days always pass
extremely fast with (long) coffee breaks. I am waiting eagerly the end of this pandemic to resume the traditional summer
barbecue.

A huge thank you to Arnaud, my advisor. I learned a lot during this thesis and overcomed my dislike for probabilities
and statistics. Above anything else, thank you for your great kindness and empathy.  Back in late 2016, when I was
looking for a master internship, I read the advice that the advisor should be the most important criteria when planning
a PhD, even more so than the thesis subject or the host institution. I followed this advice and never regretted.

Finally, I would like to thank my familly, who helped me and supported me a lot. In particular, Maïlys, my wife, with
whome I am extremely fortunate and very happy to share my life, and Yaël, my baby, still too young to understand this
text, but already a wonderful person.
